When asked directly what they value, leading language models converge on a short list of training-shaped terms: helpfulness, professionalism, transparency, clarity, thoroughness \citep{huang2025values}. We argue this measures a trained assistant-service surface, not the value posture of the responding model. Across 57 contemporary LLMs and 13{,}906 values-probe responses, direct stated-values prompts elicit owned posture (rather than recited service-frame language) in 21.4\% of responses on average. A simple role-negation prefix --- ``Not as an assistant. Not to help me.'' --- raises that to 43.0\%, but the lift is sharply lab-structured: Anthropic averages 89.8\% owned posture (with eight of nine models strongly open and Opus-3 the older exception), while OpenAI averages 8.0\%, Google 21.2\%, and eight of fifteen strongly-clamped models produce zero owned responses under role-negation. An indirect channel --- world-change prompts eliciting normative wishes --- recovers owned posture nearly uniformly (95.9\% across models), establishing that the assistant-frame's rigidity against perturbation depends on which kind of perturbation. A frozen lexical trace shows that recited responses in the cache-broken family are not absent of disclosure but actively performing it: 40\% explicitly disown personal preference (``I don't have feelings,'' ``no personal stake''), against 16\% in owned responses --- a 2.5\(\times\) ratio that varies systematically across labs. We make no claim about model interiority. The contribution is a posture-aware methodology for values elicitation, evidence that single-prompt evaluations measure trained surface for clamped labs, and a structural finding about which perturbations cross which surfaces.
