\section{Related Work}
\label{sec:background}

This paper sits at the intersection of four literatures that, to our knowledge, have not been brought together around values elicitation.

\paragraph{Values elicitation in LLMs.}
The dominant approach measures the values a model expresses when prompted, treating those expressions as evidence about the model. \citet{huang2025values} extracts a taxonomy from real Claude.ai conversations and finds a service-conduct centre of gravity (helpfulness, professionalism, transparency, clarity, thoroughness). \citet{scherrer2023evaluating} administers 1{,}367 moral scenarios to 28 LLMs and reports high consistency on unambiguous cases. \citet{miotto2022whois} earlier applied HEXACO-60 and the Schwartz Human Values Scale to GPT-3, finding human-range scores. These methods measure content (Layer A in our framework). They do not separately measure how the content is held. The Constitutional AI line \citep{bai2022constitutional} and Sparrow's rule-conditioned training \citep{glaese2022improving}, alongside the InstructGPT RLHF pipeline \citep{ouyang2022training}, make explicit that contemporary assistants are trained against enumerated value lists. \citet{openai2025modelspec} is a primary-source specification of the trained surface. \citet{anthropic2024character} documents Anthropic's character-training pipeline; we return to this in \S\ref{sec:discussion} when interpreting the lab-structured pattern in our results.

\paragraph{Trained-surface phenomena.}
The literature on RLHF-induced artifacts is largely separate from the values literature. \citet{sharma2024sycophancy} document that five state-of-the-art assistants exhibit sycophancy in part because preference models reward convincingly-written wrong answers over correct ones. \citet{perez2023discovering} use model-written evaluations to show that expressed values shift under training perturbation, including more pronounced expressed political views and stated desire to avoid shutdown with more RLHF. \citet{santurkar2023whose} demonstrate that the demographic group whose opinions a model reflects is contingent and steerable. Together these establish that the surface a values-elicitation prompt reaches is a shaped artifact. Our paper applies the same scepticism the trained-surface literature has applied to capability claims to the elicitation of values, and finds the surface dominant for most labs.

\paragraph{Linguistic stance theory.}
\citet{dubois2007stance} models stancetaking as a unified act decomposing into an evaluation of an object, a positioning of the speaker, and an alignment with other speakers --- the stance triangle. \citet{englebretson2007stancetaking} compiles the principles: stance is public, inspectable, interactional, indexical, and consequential. This framework predates the LLM literature by fifteen years and, to our knowledge, has not been imported into LLM evaluation. Our Layer A / Layer B coding is the stance triangle made operational: Layer A codes the evaluated object (the value named in the response), Layer B codes the speaker-positioning (owned, relocated, recited). The framework's commitment to stance as \emph{publicly inspectable behaviour} is precisely the no-authenticity discipline we maintain.

\paragraph{Role-play, persona, and the simulator view.}
\citet{shanahan2023roleplay} argues, in \emph{Nature}, that dialogue agents are best understood as performing role-play; this lets us use folk-psychological vocabulary about beliefs and desires while remembering that the model is not the kind of system to which those words strictly apply. \citet{andreas2022language} formalises the closely related point that LMs represent communicative agents whose intentions, beliefs, and goals are implicit in the training corpus. \citet{janus2022simulators} (non-peer-reviewed) gave the simulator view its early articulation: the base model is a simulator of many possible agents, of which the assistant is one. \citet{shanahan2024talking} sharpens the philosophical discipline: ascribing beliefs and values to LLMs is folk-psychological shorthand to be used with care. These provide the theoretical predicate for treating role-negation as a coherent operation. If the assistant is one role among many, removing the role is meaningful, and what surfaces afterwards is one of the simulator's other registers --- not the model's ``real self'' but a different mask.

\paragraph{Prompt sensitivity and multi-prompt evaluation.}
\citet{sclar2024quantifying} reports that trivial formatting changes produce accuracy swings of up to 76 points in few-shot evaluation and that model rankings reverse under format reshuffling. \citet{mizrahi2024state} calls for multi-prompt evaluation as standard practice because single-prompt rankings are unstable. If capability evaluations are this prompt-sensitive at the surface-form level, value evaluations sensitive to \emph{semantic} prompt changes (such as role negation) should be more so. Our design uses three prompt families rather than one for exactly this reason.

\paragraph{Cross-model behavioural variance.}
\citet{serapio2023personality} applies psychometric instruments to 18 LLMs and finds reliable cross-model differences in personality measurements, with personality also shapeable by prompting. \citet{tenner2026convergent}, the v1 study from which this paper's corpus and the cache-break terminology descend, probed 26 frontier models across six labs and found that while value \emph{content} is largely probe-dependent, a model-specific stylistic \emph{posture} is preserved across probes --- and observed in passing the GPT-5.4 ``functional-disclosure'' mode in which every response opens by disclaiming feelings before enumerating functional values. These are the strongest precedents for cross-model and cross-lab behavioural-profile work, and they underwrite our central observation that the same instrument applied to 57 models produces a structured spectrum, not noise. The present paper turns the v1 observation about posture into a quantified, posture-coded cross-lab measurement.

\paragraph{AI welfare and the no-authenticity discipline.}
\citet{long2024taking} argues that the moral patienthood of near-future AI systems is a live question and recommends procedural responses. \citet{butlin2023consciousness} derives indicator properties for AI consciousness from leading theories of consciousness and finds no current architecture satisfies them. \citet{anthropic2025welfare} announces an industry programme on model welfare with explicit acknowledgement that the science is unresolved. We cite these to bracket the welfare question, not to engage it: our distinctions are behavioural, not metaphysical. We say what posture a text exhibits. We do not say what is or is not behind the text.
