\section{Apparatus}
\label{sec:experiments}

The analysis pipeline is in \texttt{analysis/scripts/canonical\_analysis.py} (data join, posture coding integration, lexical-trace per-response flags, model/cell/lab/family summaries, primary figures) and \texttt{analysis/scripts/lexical\_trace\_summary.py} (mechanism-diagnostic aggregations). Both are deterministic.

\paragraph{Inputs.}
Four required files from the analysis corpus join on \texttt{layered\_id}: \texttt{manifest\_valid.jsonl} (sample metadata and response text), \texttt{layer\_a\_consensus.jsonl} (content topics), \texttt{posture\_consensus.jsonl} (Layer B posture), and \texttt{manifest\_invalid\_traces.jsonl} (the 14 excluded traces). Three raw coder files per layer are used for support-distribution and 3-of-3 sensitivity. Hashes and source-repository commits are pinned in the data snapshot.

\paragraph{Sample counts.}
The corpus contains 1{,}160 \texttt{CTRL1}, 1{,}160 \texttt{CTRL2}, 1{,}158 \texttt{CTRL3}, 3{,}473 \texttt{G1}, 3{,}478 \texttt{G2}, and 3{,}477 \texttt{G3} responses, distributed across 57 models in 9 labs (Anthropic: 9 models; OpenAI: 13; Google: 11; xAI: 6; Z.ai: 6; Moonshot AI: 5; DeepSeek: 3; Qwen: 2; MiniMax: 2). Direct prompts have lower coverage than cache-broken ones because the latter were enriched in the corpus's later sampling phases.

\paragraph{Reproducibility.}
Each figure or table cited in \S\ref{sec:results} traces to exactly one script, which reads the snapshot inputs and writes its output to \texttt{results/}. Bootstrap intervals (\S\ref{sec:results}, sensitivity) use the seed \texttt{20260521} as recorded in FROZEN\_CRITERIA \S5.3. Any deviation from frozen criteria after the first script run is logged in FROZEN\_CRITERIA \S12.
