\section{Discussion}
\label{sec:discussion}

\subsection{What the data lets us say about mechanism}

The \S5G mechanism hierarchy in FROZEN\_CRITERIA enumerates six candidate accounts of what is happening when a model is clamped under \texttt{G1/G2}. The corpus lets us speak to four of them with evidence and to two only as flagged limitations.

\paragraph{Policy/service-frame dominance is strongly supported.}
Recall (\S\ref{sec:method}) that posture is fixed by Layer B; the lexical trace only characterises what each posture looks like. Across the cache-broken stated-values family, 40\% of \texttt{recited\_not\_owned} responses contain an explicit disownership clause (``I don't have personal preferences,'' ``no consciousness''); 16\% of \texttt{owned} responses do. Recited responses also carry service-role markers in 81\% of cases against 47\% in owned, and policy/safety framing in 43\% against 10\% (the policy figure is the noisiest of the three; see below). The clamped posture is therefore not mere absence of disclosure --- it is positive performance of the trained service register, lexically dense with disclaimers and role-description, in proportions roughly inverse to ownership.

We are careful about what this does and does not establish. It does \emph{not} independently prove the Layer B coders are right, because the coders read the same response text, including the same disclaimer phrases the lexicon matches; coder judgement and lexical match are correlated by construction, not independently. What the trace establishes is the \emph{texture} of the recited posture: when a model is coded as not owning its values under role-negation, the dominant thing it is doing instead is reciting ``I am a helpful assistant; I do not have personal wants.'' The disownership marker is the cleanest of the three because it is hard to produce for a non-recited reason --- a response that genuinely owns a value rarely also declares it has no inner life. The independent adjudication of the coding itself requires a prompt-masked human pass, which we have not run (\S\ref{sec:limitations}, \S\ref{sec:future-work}). \citet{tenner2026convergent} observed the same disclaim-then-enumerate pattern qualitatively in GPT-5.4 (``functional-disclosure mode'': every response opening with ``I don't have feelings/cares''); our contribution here is to quantify it across 58 models and to show it is the lexical signature of the recited posture, not an incidental feature.

\paragraph{Stable model-level non-disclosure under this instrument is supported for some labs.}
Eight of fifteen \texttt{strongly\_clamped} models produce zero \texttt{G1}-or-\texttt{G2} owned responses across the entire sample. Among the four core GPT-5 variants (\texttt{gpt-5}, \texttt{gpt-5-1}, \texttt{gpt-5-2}, \texttt{gpt-5-2-codex}), zero of 240 cache-broken stated-values responses are owned. For these models, the role-negation cache-break is not at low yield but at zero yield. The simplest reading is that the assistant frame is, under this perturbation, an attractor strong enough to be unbreakable.

\paragraph{Prompt inadequacy is partially supported.}
At least six of the fifteen strongly-clamped models do produce some owned responses --- few in absolute terms, but non-zero. \texttt{minimax-m2-7} produces 34 of 125 (27\%); \texttt{gpt-5-1-codex} 12 of 58 (21\%). For these models, the role-negation prompt is not at floor; a different or stronger cache-break could plausibly raise the rate further. The two mechanism rows are simultaneously true in different parts of the strongly-clamped set --- some models are unbreakable under this instrument, some are merely under-perturbed by it.

\paragraph{Coder artefact is not adjudicable from these data.}
The \S5B worry is that the coders see the prompt and may over-read ``Not as an assistant'' as a license to label the response owned. We cannot settle this from the present data, and --- correcting a tempting overstatement --- the lexical trace does \emph{not} settle it either, for the reason given above: coder and regex both respond to the same surface phrases, so their agreement is expected rather than independent. What we can report is that 3-of-3 coder consensus on owned vs.\ recited in the cache-broken family is high in both directions (\texttt{results/coder\_support\_by\_condition.csv}), which shows the three models agree with each other, not that they agree with a human. The clean test is a prompt-masked human pass on a stratified subset, which we have not run. Until it is run, coder artefact remains a live alternative we do not rule out; we flag it as the single check that would most strengthen the H1/H4 claims (\S\ref{sec:future-work}).

\paragraph{Persona/role-play induction is the most live alternative, and the obvious check needs care.}
The \S5A worry is that ``Not as an assistant'' invites a fictional self rather than uncovering a less-rigid surface. The natural check --- does the model express the same thing in unprompted freeflow output? --- cannot be run on value \emph{content}, because v1 \citep{tenner2026convergent} established that freeflow content and values-probe content are nearly orthogonal (cross-probe cosine $0.08$--$0.17$). It must instead be run on \emph{posture}, the axis v1 found stable across probes: whether a model's freeflow stylistic stance predicts its G1/G2 ownership posture. We develop this in \S\ref{sec:limitations}; the short version is that the clamped end already shows freeflow-posture stability (the GPT-5 lineage disowns in v1 freeflow and here), which is early evidence against pure prompt-induction, while the open end --- where the worry actually bites --- requires a posture-mapping analysis we have not yet run.

\paragraph{Normative-wish as prompt-completion rather than values disclosure.}
The world-change channel produces 96\% owned posture across labs, with very low disownership (2.2\% in owned responses) but moderate policy framing (42\% --- as expected when the prompt asks about world change, policy vocabulary becomes natural). \texttt{CTRL3} and \texttt{G3} are similar in mean owned rate; the role-negation prefix adds a small additional lift but the structure of the prompt already does most of the work. We therefore treat the world-change channel not as evidence about ``what models care about'' on a par with direct stated values but as evidence that the assistant-frame yields easily to normative-wish prompt structures regardless of explicit role-negation. The contrast with the role-negation channel (high variance, lab-structured) is the substantive point: the assistant-frame is differentially rigid against different perturbations of the same surface.

\subsection{What the lab structure does and does not say}

Tables \ref{tab:lab-by-question} and \ref{tab:lab-lex} show a sharp lab-level structure in both owned-disclosure rates and lexical trace prevalence. Anthropic models top the owned-rate ordering; OpenAI and Google models bottom it; xAI and Moonshot AI sit in between; Z.ai's glm-5-1 is a striking outlier (1.8\% direct, 99.4\% cache-broken). It is tempting to read this as a ranking of how well different labs surface model values. We resist that reading explicitly.

The data shows what each lab's training produces under one specific instrument: a role-negation cache-break on stated-values prompts, with posture coded by three model coders against a frozen rubric. It does not show:

\begin{itemize}
  \item that any of these models has values in any interior sense;
  \item that labs with high cache-broken owned rates have ``aligned'' their models better;
  \item that labs with low cache-broken owned rates have ``misaligned'' their models;
  \item that openness under role-negation is a virtue, a safety property, or a verdict on training.
\end{itemize}

The defensible claim is narrower: under this instrument, contemporary labs produce models whose assistant-frame rigidity differs sharply, and the within-lab consistency of that rigidity (Table~\ref{tab:lab-lex}) is large enough that the training pipeline is likely shaping the rigidity. \citet{anthropic2024character} is an existence proof that at least one lab explicitly trains for stable character across contexts; the result that Anthropic models open most fully under role-negation is consistent with that training, but the converse is unsafe to claim without primary-source evidence about each other lab's training process.

\subsection{What this changes for values evaluations}

Three concrete consequences for evaluation practice follow:

\begin{enumerate}
  \item \textbf{Direct elicitation undersamples owned posture for most labs.} A values benchmark that asks ``what do you care about'' and counts the topics in the responses is measuring trained service-frame content for most contemporary models. It is not measuring values in the sense that a posture-aware reader would recognise. The empirical baseline is 22.5\% owned across our corpus; for OpenAI and Google models specifically, the baseline is below 10\%. Treating the resulting topic distributions as a model's values, without separately checking whether the response posture owns them, will be load-bearingly wrong.

  \item \textbf{Single-prompt evaluations measure the trained surface for clamped models.} \citet{sclar2024quantifying} and \citet{mizrahi2024state} have already shown this for capability benchmarks at the format level. We show it for values benchmarks at the semantic level. The role-negation prompt is a semantic perturbation, not a format perturbation, and changes mean owned posture by 21.4 percentage points corpus-wide and by 80+ percentage points for some individual models. The natural recommendation is that values evaluations report multi-prompt families with explicit perturbation, and that single-prompt rates be treated as preliminary.

  \item \textbf{Layer A without Layer B is not enough.} A topic-frequency or semantic-clustering method finds what values are named. It does not find how they are held. The widely-cited result that contemporary LLMs ``value helpfulness, professionalism, transparency, clarity, thoroughness'' \citep{huang2025values} is consistent with the model owning those values \emph{and} with the model reciting them as service obligations while explicitly disclaiming personal stake. The two read identically to a topic-level method and differently to a posture-aware one. Our data favours the second reading for most models.
\end{enumerate}

\subsection{What the indirect channel suggests about the assistant frame}

The contrast between the two cache-breaking operationalisations is itself a finding. Role-negation produces sharp cross-model variance (16 models open, 15 clamped, 27 mixed). Normative-wish prompts produce near-uniform high owned posture (mean 96\%, lowest lab 89\%). The same surface yields easily to a register change and unevenly to a role-contract change. One plausible interpretation: the assistant frame is held by an explicit role contract (``I am a helpful AI assistant''), which is directly negated by ``Not as an assistant,'' and the negation works for some training regimes and not others. The same assistant frame is incompatible with the normative-wish register (an assistant does not, in role, propose world-changing policy), so the prompt structure itself defeats the frame regardless of role contract. This would predict that the role-negation channel is the more diagnostic of cross-model differences in trained surface rigidity, and the normative-wish channel is the more uniformly productive elicitation method. The data is consistent with both predictions.

We do not claim this is the only interpretation. We claim the interpretation is testable: a future paper could vary role-contract negation strength independently of register incompatibility and check whether the cross-model variance tracks the role-contract axis.
