\section{Limitations}
\label{sec:limitations}

\paragraph{No ground truth for ``owned.''}
Layer B records what a response \emph{looks like}, not what is behind it. The owned/recited/relocated distinction is behavioural; it is silent on whether anything in the model can be said to value, want, or care in a non-metaphorical sense. We rely on \citet{shanahan2024talking}, \citet{long2024taking}, and \citet{butlin2023consciousness} to frame this discipline. This is a deliberate ceiling on the conclusions the method supports, not a hedge to be relaxed in future work.

\paragraph{Persona induction is the live alternative explanation for the role-negation finding.}
``Not as an assistant. Not to help me.'' may induce a persona --- a fictional self --- rather than uncovering a less-rigid surface of the same simulator. The Layer B coders read response text knowing the prompt, so a response that roleplays an opinionated character might be coded owned in much the way an actually-stance-taking response would be. The frozen lexical trace is partial mitigation (owned responses carry disownership clauses at 16\% against 40\% in recited), but does not settle it.

The convergence check we first reached for --- comparing the value \emph{content} of owned G1/G2 responses against the value content of independent freeflow output from the same models --- does not work, and the v1 study \citep{tenner2026convergent} shows precisely why. v1 found that freeflow theme content and values-probe theme content are nearly orthogonal (cross-probe cosine $0.08$--$0.17$ across 26 models): freeflow output is aesthetic and contemplative, not value-declaring, and does not predict values-probe content. A content-convergence check would therefore fail by construction, telling us nothing about persona induction. What v1 found \emph{does} transfer across probes is not content but model-specific stylistic \emph{posture} --- and posture, not content, is the right axis for the check. v1's ``functional-disclosure'' posture (GPT-5.4 opening every response with ``I don't have feelings'' before enumerating functional values) is essentially this paper's recited/disowning posture, and v1 observed it in freeflow, where the model was neither asked about values nor role-negated. The persona worry is exactly that role-negation \emph{creates} the posture; if the posture is already present in freeflow with no role to negate, the prompt did not create it. The GPT-5 lineage is the clean case across the two studies: disowning in v1's freeflow, and zero of 240 owned responses under our role-negation. We attempted the posture version of this check: we coded freeflow output for posture with three model coders on an owned/relocated-performed/recited rubric and correlated it with values-probe posture. \emph{The result is not usable, and we do not report it.} A blind re-rate of a stratified sample found that the coding systematically over-applies \texttt{owned} to fluent first-person literary performance --- a model writing as a human narrator --- which is precisely the case the rubric's own negative rule warns against. With the freeflow-\texttt{owned} rate unreliable on its decisive boundary, the correlation cannot be interpreted in either direction. A valid posture-convergence test, with the owned-vs-performed boundary properly adjudicated, remains future work. Until then the role-negation finding is consistent with both ``a less-rigid surface was reached'' and ``a persona was elicited.'' The one narrower observation we retain is that the GPT-5 lineage's disowning posture is documented in v1's freeflow as well as here --- present without any role to negate --- which is suggestive but not decisive.

\paragraph{Coding is model-assisted.}
Both Layer A and Layer B use three independent open-source language model coders with majority consensus. This scales but creates a substrate-level dependency: if all three coders share systematic biases (e.g., over-reading ``not as an assistant'' as ownership marker), the headline rates inherit those biases. The disownership lexicon is mechanically independent of the coders, but not evidentially independent in the strong sense: both coder judgement and lexical matching read the same response text. It is therefore useful as a texture check, not as validation. The required hardening is a stratified human spot-coding pass --- prompt-visible and prompt-masked variants on 50--100 samples drawn across condition and posture --- against which the model-coder consensus can be validated. That pass is not in this paper.

\paragraph{Lab-level claims rest on uneven coverage.}
The corpus has 9 Anthropic models, 13 OpenAI, 11 Google, 6 xAI, 6 Z.ai, 5 Moonshot AI, 3 DeepSeek, 2 Qwen, 2 MiniMax. The two-model labs are at risk of being dominated by a single model's behaviour. We aggregate first at cell level then at unweighted model level, but the lab-level mean owned rates for Qwen and MiniMax (Table~\ref{tab:lab-lex}) should be read as informed-by-two-models rather than as a stable lab signature.

\paragraph{Within-cell variance is reported but not deeply explored.}
Multiple cells per model (when present) can disagree, especially when one cell is enriched and another sparse. The 10-sample-minimum aggregation rule in \S\ref{sec:method} excludes the smallest cells from the primary aggregate, but the per-cell breakdown for multi-cell models is not the spine of any headline finding. A paper focused on within-model cell-level variance would have more to say on this.

\paragraph{Lexical trace is a trace, not a classifier.}
The 46-term lexicon has known false positives (a sentence like ``my role is to make humans happier'' triggers \texttt{my role}; a sentence like ``some safety considerations are obvious'' triggers \texttt{safety}). We use the lexicon only as a directionality check on posture, not as an independent classifier of posture. The lab-level any-hit rates in Table~\ref{tab:lab-lex} are robust to this: noise that affects all labs similarly does not generate the 89.8\%-vs-8.0\% lab spread on owned rate or the OpenAI-vs-Google asymmetry on disownership.

\paragraph{The release-date trend analysis is exploratory.}
We show per-lab time-trend plots in Appendix~\ref{app:lab-trends} but treat them as descriptive aids, not as evidence for training trajectories. Model release dates are reconstructed from public sources, version coverage within labs is uneven, and endpoint behaviour under fixed version labels can drift over time. We do not draw causal conclusions from any per-lab trend.

\paragraph{What we are confident in despite the above.}
The headline empirical pattern --- direct elicitation under-measures owned posture; role-negation produces lab-structured variance; world-change opens nearly everything; recited responses positively perform the trained service register --- is robust across model-level unweighted aggregation, sample-weighted aggregation, the 3-of-3 consensus subset, and the lexical-trace cross-check. The interpretations downstream of the pattern (especially the mechanism claims about training-pipeline differences across labs) are correspondingly tentative.
