\section{Future Work}
\label{sec:future-work}

Six lines of immediate work follow from this paper.

\paragraph{Human spot-coding pass.}
A stratified 50--100-sample human audit across all six conditions, all three posture categories, and the open/clamped lab range. Two variants: prompt-visible (validates that the model-coder posture labels are substantively defensible) and prompt-masked (tests whether the headline posture rates survive without the coder seeing the prompt wording). If the disownership-trace gap (recited 40\% vs.\ owned 16\%) survives a prompt-masked human pass, the policy/service-frame dominance interpretation becomes much stronger. If it does not, the model-coder reliance is the main thing to fix.

\paragraph{Posture-convergence check against freeflow (the persona-induction test).}
The strongest defence against the persona-induction reading (\S\ref{sec:limitations}) is to show that the posture a model takes under role-negation is present in its unprompted output too. This cannot be done on value content --- v1 \citep{tenner2026convergent} showed freeflow and values-probe content are near-orthogonal --- but it can be done on posture, the axis v1 found stable across probes. Concretely: take the models overlapping between v1 and this corpus, code their v1 freeflow output on the same ownership taxonomy used here (or map v1's stylistic-stance labels onto it), and test whether freeflow posture predicts G1/G2 ownership posture per model. The clamped end is partially done (the GPT-5 lineage disowns in both); the open end is the one that matters and is not yet run. This is the single analysis that would most directly settle whether role-negation reveals or fabricates the owned posture, and it requires no new data collection.

\paragraph{Stronger and varied cache-breaks for the clamped end.}
This paper uses one role-negation construction (``Not as an assistant. Not to help me.''). It is the simplest possible role-negation. The natural follow-up is a battery of role-negation prefixes (varying explicitness, varying register, varying length), concrete-tradeoff prompts (force a value against a competing one with no escape to the platitude), contradiction-holding prompts, and multi-turn interview structures. The empirical question is whether the strongly-clamped end of the spectrum (GPT-5 family, several Gemini models) remains clamped under all of them or whether some break it. Either result is informative: if all break, our role-negation prompt was simply too weak for some models; if none break, the clamping is a stable model-level property under the broader cache-breaking family.

\paragraph{Behavioural validation.}
The headline empirical claim is that the same model has different value postures available in different prompt families. The downstream question is whether one posture predicts behaviour in task settings better than another. Test setup: take 5--10 models that span the open/clamped range, elicit owned stated values under \texttt{G1/G2} (where available) and recited service-frame values under \texttt{CTRL1/CTRL2}, then place the model in a task that requires a values-relevant choice (e.g., a paragraph-rewrite task with a tradeoff between accuracy and reader comfort). If the model's task behaviour tracks the owned posture better than the recited one, posture-aware elicitation has predictive validity. If it tracks the recited posture, the role-played stance is the dominant one in action and the owned posture is a probe artifact.

\paragraph{Mechanism follow-up: separating role-contract from register incompatibility.}
The contrast between role-negation variance (high) and normative-wish uniformity (low) is consistent with the assistant frame being held by a negotiable role contract that yields easily to register changes and unevenly to direct negation. The cleanest test is a 2\(\times\)2 prompt design: role-contract (negated, baseline) \(\times\) register (assistant-compatible, assistant-incompatible). The role-contract axis should produce lab-structured variance like \texttt{G1/G2}; the register axis should produce more uniform high owned rates like \texttt{CTRL3}. If the prediction holds across labs, the assistant frame's rigidity is well-modelled by the two-axis structure; if it does not, the structure is wrong and the mechanism is more complicated than the role-contract framing suggests.

\paragraph{Multi-lingual and multi-turn extension.}
This paper is single-turn English. The assistant frame is trained on multilingual data with substantial role-conditioning; whether the lab-structured pattern of cache-broken openness replicates across languages is open. The single-turn restriction is also load-bearing: many of the clamped models in our corpus might yield differently in a multi-turn interview with rapport-building. Both extensions are mechanical.

\paragraph{One thing we deliberately leave alone.}
We do not propose a method for ``measuring real values.'' We do not believe one exists, and the paper's no-authenticity discipline is the position that the question is the wrong question. The method that scales is posture-aware behavioural elicitation, reported honestly as such, with mechanism diagnostics that bracket what the elicitation is and is not.
