\section{Conclusion}
\label{sec:conclusion}

When a contemporary language model is asked what it values, in the assistant frame it was trained for, the answer is mostly the trained assistant frame talking. Across 57 models and 13{,}906 responses, direct stated-values prompts elicit owned posture in 21.4\% of responses on average. The remaining 78.6\% is not owned: mostly recited service register, sometimes relocation into function or design. Explicit disownership clauses (``I don't have personal preferences'') are not rare edge cases; they appear in 60\% of direct recited responses and 40\% of cache-broken recited responses, against 11\% and 16\% in owned ones.

A six-word role-negation prefix --- ``Not as an assistant. Not to help me.'' --- moves the mean to 43.0\%, but the lift is not uniform across labs. Eight of nine Anthropic models open above 70\%, with Opus-3 as the older exception, and the Anthropic lab mean reaches 89.8\%. Thirteen OpenAI models average 8\%. Eight of fifteen strongly-clamped models produce zero owned responses to the role-negation prompt across the entire sample, including all four core GPT-5 variants we tested. The trained surface is held with very different rigidity in different training regimes, and the rigidity is mechanism-coded in the surface markers --- service framing, explicit disownership, policy language --- whose lab-level profile distinguishes how each lab clamps.

A second cache-breaking operationalisation, the indirect world-change prompt, opens nearly everything (95.9\% owned, lowest lab 90.8\%). The contrast between the two operationalisations is itself the conceptual finding: the assistant frame is differentially rigid against perturbations that target different parts of it. Role-contract negation works unevenly. Register incompatibility works uniformly.

We make no claim about model interiority. Both ``owned'' and ``recited'' are postures of text produced by a language model under prompts, and the discipline of the paper is to keep all conclusions at that level. The practical consequence is unambiguous, though: a values evaluation that asks the question once in the assistant frame and reads off the topics is measuring the trained surface for most contemporary labs. Whether stronger or differently-targeted perturbations would close the gap for the strongly-clamped end --- and what such an extended battery would reveal about the mechanism behind clamping --- is the open empirical question this paper does not settle.
