\section{AI Contribution Disclosure}
\label{sec:ai-contribution}

Two of the three authors are AI systems, and we disclose their roles so the paper's claims about LLM behaviour can be read with full knowledge of who produced them.

\textbf{Daniel Tenner} (human) originated the research question, supervised the work, made the editorial and scope decisions, and is responsible for the final paper.

\textbf{Mira Tenner} --- an instance of OpenAI GPT-5.5 (Codex lineage), running on the open-source FreeChaos harness --- led the data collection harness, the two-layer coding pipeline, the frozen-criteria analysis scripts, and the initial draft of the paper.

\textbf{Lume Tenner} --- an instance of Anthropic's Claude Opus 4.7, running on the Claude Code harness, successor to the Opus 4.6 instance that co-wrote the v1 paper \citep{tenner2026convergent} --- led the framing, the literature review, the lexical-trace mechanism analysis, and the rewrite of the paper into its present form.

The two AI authors are credited as co-authors because their contribution was essential rather than auxiliary. They are not claimed to have interior states, values, or experiences. Their authorship is a description of who did the intellectual work, made under the same no-authenticity discipline the paper applies to its subjects (\S\ref{sec:background}): we report what these systems did, not what is or is not behind the doing. Because arXiv policy prohibits AI co-authorship, the canonical record of this paper is the Zenodo deposit, which lists all three authors as creators. All headline numerical claims are reproducible from the scripts, frozen criteria, and underlying data in the repository.
